---
title: "Generation interval from contact tracing"
format:
  html:
    code-fold: true
    code-summary: "show code"
    embed-resources: true
execute:
  warning: false
  message: false
---

Both ends of a generation interval are infection times, and here each is known
only to within a contact window. So both ends are interval censored: the
infector's infection (primary) and the infectee's (secondary). `primarycensored`
fits the delay by integrating over both windows. Using infection times rather
than onsets makes this a generation interval, not a serial interval.

The pairs come from tracing back from cases to their infectors, so it is a
backward sample. In a growing epidemic that contracts the interval: recent
infectors are more numerous, so the gap to them looks short. Weighting each pair
by `exp(r·τ)` undoes the tilt and recovers a forward, roughly intrinsic interval.

```{r}
library(dplyr)
library(primarycensored)
library(ggplot2)

ct    <- readRDS(here::here("data", "contact_tracing_data.rds"))
inc40 <- readRDS(here::here("data", "incidence_data_day40.rds"))

# keep pairs where the infector was also traced, so both infection windows exist
windows <- ct |> distinct(ind_id, .keep_all = TRUE) |>
  select(ind_id, lower = time_of_contact_lower, upper = time_of_contact_upper)
gi <- ct |>
  select(infector = reported_contact,
         inf_lower = time_of_contact_lower, inf_upper = time_of_contact_upper) |>
  inner_join(windows, by = c("infector" = "ind_id")) |>
  mutate(pwindow = (upper - lower) + 1,         # infector window (primary)
         left    = pmax(inf_lower - lower, 0),  # infectee window (secondary)
         right   = (inf_upper + 1) - lower,
         swindow = right - pmax(inf_lower - lower, 0),
         tau_mid = ((inf_lower + inf_upper) / 2) - ((lower + upper) / 2),
         .keep = "none")

# doubly interval-censored gamma; weights turn the backward fit into a forward one
fit <- function(w) {
  nll <- function(p) -sum(w / mean(w) * vapply(seq_len(nrow(gi)), \(i)
    dprimarycensored(gi$left[i], \(q) pgamma(q, exp(p[1]), exp(p[2])),
      pwindow = gi$pwindow[i], swindow = gi$swindow[i], log = TRUE), numeric(1)))
  o <- optim(c(log(2), log(.5)), nll, hessian = TRUE, method = "BFGS")
  list(par = o$par, vcov = solve(o$hessian))
}
# mean and sd, with CIs from sampling the (log) parameters
set.seed(1)
ci <- function(f) {
  d <- exp(sweep(matrix(rnorm(8000), ncol = 2) %*% chol(f$vcov), 2, f$par, "+"))
  list(mean = d[, 1] / d[, 2], sd = sqrt(d[, 1]) / d[, 2])
}
```

Growth comes from the rising phase of the onset curve, as the slope of a
quasi-Poisson GLM (quasi-Poisson because the daily counts are badly overdispersed).

```{r}
m  <- glm(n ~ time_of_symptom_onset, quasipoisson,
          count(filter(inc40, between(time_of_symptom_onset, 8, 26)), time_of_symptom_onset))
r  <- unname(coef(m)[2]); r_se <- sqrt(vcov(m)[2, 2])
sprintf("growth rate r = %.2f /day (95%% CI %.2f, %.2f)", r, r - 1.96 * r_se, r + 1.96 * r_se)
```

```{r}
draws <- list(
  "backward (observed)"       = ci(fit(rep(1, nrow(gi)))),
  "forward (growth-weighted)" = ci(fit(exp(r * gi$tau_mid)))
)
fmt <- \(x) sprintf("%.1f (%.1f, %.1f)", median(x), quantile(x, .025), quantile(x, .975))
purrr::imap_dfr(draws, \(d, nm) tibble::tibble(
  estimand = nm, `mean (days)` = fmt(d$mean), `sd (days)` = fmt(d$sd))) |>
  knitr::kable()
```

The backward interval is about 3 days; undoing the tilt moves it to about 5. `r`
is tight enough (above) that propagating its uncertainty barely changes the
forward CI, so the band is the fit uncertainty at the point `r`, and most of the
width is the 68 pairs.

## Joint Bayesian model with forward correction

The growth-weighting above removes the backward bias indirectly through the
growth rate. A better option is to remove it directly. We fit a joint BAyesian model over three
streams of the same data: the incubation period from every traced infectee, the
chain generation intervals where both infection windows are known, and the onset
generation intervals where only the infector's onset is known and its incubation
is integrated out. It right-truncates each pair at the tracing cutoff. The generation interval and the incubation period are
both estimated as parametric gamma distributions, and the mean and uncertainty of each is the parameter reported below. The code is in
`analysis/joint_model/joint.jl`.

Sampling takes about 25 minutes on a Macbook, so it runs separately and writes the draws to
CSV, which we read here.

```{r}
jdraws <- read.csv(here::here("analysis", "joint_model", "joint_gi_draws.csv"))

fmt <- \(x) sprintf("%.2f (%.2f, %.2f)", median(x), quantile(x, .025), quantile(x, .975))
tibble::tibble(
  estimand      = c("generation interval", "incubation period"),
  `mean (days)` = c(fmt(jdraws$mu_g), fmt(jdraws$mu_i)),
  `sd (days)`   = c(fmt(jdraws$mu_g / sqrt(jdraws$k_g)),
                    fmt(jdraws$mu_i / sqrt(jdraws$k_i)))) |>
  knitr::kable()
```

All three generation-interval estimates on one set of axes, over the observed
backward delays (grey): the backward fit, the growth-weighted forward fit, and the
joint model's forward-corrected fit.

```{r}
#| fig-width: 6
#| fig-height: 3.5
# all three GI densities with a 95% band each (gamma recovered from each fit's draws)
grid <- seq(0, 14, 0.1)
band <- function(sh, rt, est) {
  dn <- vapply(seq_along(sh), \(j) dgamma(grid, sh[j], rt[j]), numeric(length(grid)))
  tibble::tibble(t = grid, estimand = est, med = apply(dn, 1, median),
                 lo = apply(dn, 1, quantile, .025), hi = apply(dn, 1, quantile, .975))
}
# backward and growth-weighted fits (R, mean/sd draws); joint model (gamma params)
bands <- dplyr::bind_rows(
  purrr::imap_dfr(draws, \(d, est) {
    i <- sample(length(d$mean), 600)
    band((d$mean[i] / d$sd[i])^2, d$mean[i] / d$sd[i]^2, est)
  }),
  band(jdraws$k_g, jdraws$k_g / jdraws$mu_g, "forward (joint model)")
)
bands$estimand <- factor(bands$estimand,
  levels = c("backward (observed)", "forward (growth-weighted)", "forward (joint model)"))

ggplot(bands, aes(t, med, colour = estimand, fill = estimand)) +
  geom_histogram(data = tibble::tibble(d = gi$tau_mid), aes(d, after_stat(density)),
                 inherit.aes = FALSE, binwidth = 1, fill = "grey88", colour = "white") +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 1) +
  labs(x = "generation interval (days)", y = "density", colour = NULL, fill = NULL) +
  theme_minimal() + theme(legend.position = "bottom")
```
