---
title: "Household transtat data construction"
format:
  html:
    code-fold: true
    code-summary: "show code"
    embed-resources: true
execute:
  warning: false
  message: false
---

## Purpose

This file creates the household pair data, which will be used for donwnstream 
analyses with `transtat`. It is intentionally limited to data construction 
so that the generated CSV can be pushed together with this qmd.

The output is located at:

```{r}
output_csv <- here::here("analysis/group-B/household_transtat_data.csv")
tibble::tibble(output = output_csv) |>
  knitr::kable()
```

## Assumptions

```{r}
library(dplyr)
library(readr)

observation_end <- 40
assumed_infectious_duration <- 14

tibble::tribble(
  ~quantity, ~value, ~meaning,
  "`observation_end`", as.character(observation_end), "end of observed follow-up",
  "`assumed_infectious_duration`", as.character(assumed_infectious_duration), "days after symptom onset during which a household member is treated as a possible source",
  "`infectious_onset`", "symptom onset", "used as the source time origin because infection times are not observed for all household members"
) |>
  knitr::kable()
```

## Input data

```{r}
ct <- readRDS(here::here("data/contact_tracing_data.rds"))
dem <- readRDS(here::here("data/demographic_data.rds"))
inc40 <- readRDS(here::here("data/incidence_data_day40.rds"))

tibble::tribble(
  ~object, ~role,
  "`contact_tracing_data.rds`", "traced pairs, pair-specific symptom onsets, and contact-time windows",
  "`demographic_data.rds`", "household membership and individual covariates",
  "`incidence_data_day40.rds`", "observed symptom-onset times up to day 40"
) |>
  knitr::kable()
```

## Person-level onset data

For each person, symptom onset is the earliest non-missing onset found in either
the incidence data or the contact-tracing data. People without an onset by day
40 are treated as not observed infected.

```{r}
onsets <- bind_rows(
  inc40 |>
    transmute(person_id = ind_id, symptom_onset = time_of_symptom_onset),
  ct |>
    transmute(person_id = ind_id, symptom_onset = ind_symptom_onset),
  ct |>
    transmute(person_id = reported_contact, symptom_onset = reported_contact_symptom_onset)
) |>
  filter(!is.na(symptom_onset)) |>
  group_by(person_id) |>
  summarise(symptom_onset = min(symptom_onset), .groups = "drop")

people <- dem |>
  transmute(
    person_id = ind_id,
    hh_id,
    age,
    sex,
    classroom_id,
    workplace_id
  ) |>
  left_join(onsets, by = "person_id") |>
  mutate(
    infected_observed = !is.na(symptom_onset),
    infectious_onset = symptom_onset,
    infectious_end = symptom_onset + assumed_infectious_duration
  )

people |>
  summarise(
    people = n(),
    observed_infected = sum(infected_observed),
    observed_uninfected = sum(!infected_observed)
  ) |>
  knitr::kable()
```

## Household subset

The subset is restricted to households that contain at least one contact-tracing
pair where `reported_contact` and `ind_id` share the same household. Here, the 
direction of transmission is determined by the order of symptom onsets (for the 
sake of simplicity). 

```{r}
household_contact_pairs <- ct |>
  transmute(
    infector_id = reported_contact,
    infectee_id = ind_id,
    infector_symptom_onset = reported_contact_symptom_onset,
    infectee_symptom_onset = ind_symptom_onset,
    contact_lower = time_of_contact_lower,
    contact_upper = time_of_contact_upper,
    contact_mid = (time_of_contact_lower + time_of_contact_upper) / 2
  ) |>
  left_join(
    people |> select(infector_id = person_id, infector_hh_id = hh_id),
    by = "infector_id"
  ) |>
  left_join(
    people |> select(infectee_id = person_id, infectee_hh_id = hh_id),
    by = "infectee_id"
  ) |>
  mutate(same_household = infector_hh_id == infectee_hh_id) |>
  filter(same_household)

household_members <- people |>
  filter(hh_id %in% household_contact_pairs$infector_hh_id)

household_summary <- household_members |>
  group_by(hh_id) |>
  summarise(
    household_size = n(),
    n_observed_infected = sum(infected_observed),
    n_observed_uninfected = sum(!infected_observed),
    n_household_contact_pairs = sum(household_contact_pairs$infector_hh_id == first(hh_id)),
    .groups = "drop"
  )

household_summary |>
  summarise(
    households = n(),
    members = sum(household_size),
    observed_infected = sum(n_observed_infected),
    observed_uninfected = sum(n_observed_uninfected)
  ) |>
  knitr::kable()
```

## transtat rows

For each selected household, the data contain:

- internal ordered rows `i -> j` for each observed infected household member
  `i` and each other household member `j`;
- one external-source row `0 -> j` for each household member `j`.

For internal rows, `Surv(start, stop, event)` is on the onset-to-onset time
scale. `event = 1` means that the row is a possible source for the susceptible's
observed event, not that the true infector is known.

```{r}
infectors <- household_members |>
  filter(infected_observed) |>
  select(
    hh_id,
    infector = person_id,
    infector_age = age,
    infector_sex = sex,
    infector_onset = infectious_onset,
    infector_end = infectious_end
  )

susceptibles <- household_members |>
  select(
    hh_id,
    sus = person_id,
    sus_age = age,
    sus_sex = sex,
    sus_onset = symptom_onset,
    sus_infected_observed = infected_observed
  )

internal_rows <- inner_join(
  infectors,
  susceptibles,
  by = "hh_id",
  relationship = "many-to-many"
) |>
  filter(infector != sus) |>
  mutate(
    stop_calendar = if_else(sus_infected_observed, sus_onset, observation_end),
    start = 0,
    stop = pmax(0, pmin(stop_calendar - infector_onset, assumed_infectious_duration)),
    event = as.integer(
      sus_infected_observed &
        infector_onset < sus_onset &
        infector_end >= sus_onset
    ),
    external = 0L,
    pair_type = "household_internal",
    same_household = TRUE,
    observed_ct_household_pair = paste(infector, sus) %in%
      paste(household_contact_pairs$infector_id, household_contact_pairs$infectee_id)
  ) |>
  filter(stop > start) |>
  select(
    hh_id,
    sus,
    infector,
    start,
    stop,
    event,
    external,
    pair_type,
    same_household,
    observed_ct_household_pair,
    sus_infected_observed,
    sus_onset,
    infector_onset,
    infector_end,
    sus_age,
    sus_sex,
    infector_age,
    infector_sex
  )

external_rows <- susceptibles |>
  mutate(
    infector = "external",
    start = 0,
    stop = if_else(sus_infected_observed, sus_onset, observation_end),
    event = as.integer(sus_infected_observed),
    external = 1L,
    pair_type = "external",
    same_household = FALSE,
    observed_ct_household_pair = FALSE,
    infector_onset = NA_real_,
    infector_end = NA_real_,
    infector_age = NA_integer_,
    infector_sex = NA_character_
  ) |>
  filter(stop > start) |>
  select(
    hh_id,
    sus,
    infector,
    start,
    stop,
    event,
    external,
    pair_type,
    same_household,
    observed_ct_household_pair,
    sus_infected_observed,
    sus_onset,
    infector_onset,
    infector_end,
    sus_age,
    sus_sex,
    infector_age,
    infector_sex
  )

household_transtat_data <- bind_rows(internal_rows, external_rows) |>
  arrange(hh_id, sus, external, infector) |>
  mutate(row_id = row_number(), .before = 1)

household_transtat_data |>
  count(pair_type, event) |>
  knitr::kable()
```

## Write CSV

```{r}
write_csv(household_transtat_data, output_csv)

tibble::tibble(
  file = output_csv,
  rows = nrow(household_transtat_data),
  columns = ncol(household_transtat_data)
) |>
  knitr::kable()
```

## Data items

The CSV contains one row per possible source-to-susceptible pair. Internal rows
represent possible household transmission; external rows represent possible
infection from outside the observed household.

```{r}
data_item_descriptions <- tibble::tribble(
  ~column, ~description,
  "row_id", "unique row identifier in the constructed CSV",
  "hh_id", "household ID",
  "sus", "susceptible / recipient individual ID for this row",
  "infector", "possible source individual ID, or `external` for the outside-household source",
  "start", "start of the risk interval used in `Surv(start, stop, event)`; currently 0 for all rows",
  "stop", "end of the risk interval; for internal rows this is truncated at 14 days after source onset",
  "event", "1 if this row is a possible source for an observed infection event; 0 if censored",
  "external", "1 for external-source rows; 0 for household-internal rows",
  "pair_type", "`household_internal` or `external`",
  "same_household", "TRUE for internal household rows; FALSE for external rows",
  "observed_ct_household_pair", "TRUE if the ordered internal pair also appears as a same-household contact-tracing pair",
  "sus_infected_observed", "TRUE if the susceptible had an observed symptom onset by day 40",
  "sus_onset", "observed symptom onset time for the susceptible, if any",
  "infector_onset", "observed symptom onset time for the possible household source; missing for external rows",
  "infector_end", "assumed end of infectiousness for the possible household source; `infector_onset + 14`",
  "sus_age", "age of the susceptible",
  "sus_sex", "sex of the susceptible",
  "infector_age", "age of the possible household source; missing for external rows",
  "infector_sex", "sex of the possible household source; missing for external rows"
)

data_item_types <- tibble::tibble(
  column = names(household_transtat_data),
  type = vapply(household_transtat_data, \(x) paste(class(x), collapse = ", "), character(1))
)

data_item_types |>
  left_join(data_item_descriptions, by = "column") |>
  knitr::kable()
```
