---
title: "Generation interval from contact tracing"
format:
  html:
    code-fold: true
    code-summary: "show code"
    embed-resources: true
execute:
  warning: false
  message: false
---

Both ends of a generation interval are infection times, and here each is known
only to within a contact window. So both ends are interval censored: the
infector's infection (primary) and the infectee's (secondary). `primarycensored`
fits the delay by integrating over both windows. Using infection times rather
than onsets makes this a generation interval, not a serial interval.

The pairs come from tracing back from cases to their infectors, so it is a
backward sample. In a growing epidemic that contracts the interval: recent
infectors are more numerous, so the gap to them looks short. Weighting each pair
by `exp(r·τ)` undoes the tilt and recovers a forward, roughly intrinsic interval.

```{r}
library(dplyr)
library(primarycensored)
library(ggplot2)

ct <- readRDS(here::here("data", "contact_tracing_data.rds"))
inc40 <- readRDS(here::here("data", "incidence_data_day40.rds"))

# keep pairs where the infector was also traced, so both infection windows exist
windows <- ct |>
  distinct(ind_id, .keep_all = TRUE) |>
  select(ind_id, lower = time_of_contact_lower, upper = time_of_contact_upper)

gi <- ct |>
  select(
    infector = reported_contact,
    inf_lower = time_of_contact_lower,
    inf_upper = time_of_contact_upper
  ) |>
  inner_join(windows, join_by(infector == ind_id)) |>
  mutate(
    pwindow = (upper - lower) + 1, # infector window (primary)
    left = pmax(inf_lower - lower, 0), # infectee window (secondary)
    right = (inf_upper + 1) - lower,
    swindow = right - pmax(inf_lower - lower, 0),
    tau_mid = ((inf_lower + inf_upper) / 2) - ((lower + upper) / 2),
    .keep = "none"
  )

# doubly interval-censored gamma; weights turn the backward fit into a forward one
fit <- function(w) {
  w <- w / mean(w)
  nll <- function(p) {
    dens <- \(q) pgamma(q, exp(p[1]), exp(p[2]))
    ll <- mapply(
      \(l, pw, sw) dprimarycensored(l, dens, pwindow = pw, swindow = sw, log = TRUE),
      gi$left, gi$pwindow, gi$swindow
    )
    -sum(w * ll)
  }
  o <- optim(c(log(2), log(0.5)), nll, hessian = TRUE, method = "BFGS")
  list(par = o$par, vcov = solve(o$hessian))
}

# mean and sd, drawn from the fitted (log) parameter covariance
draw <- function(f) {
  d <- exp(sweep(matrix(rnorm(2 * 4000), ncol = 2) %*% chol(f$vcov), 2, f$par, "+"))
  list(mean = d[, 1] / d[, 2], sd = sqrt(d[, 1]) / d[, 2])
}
```

Growth comes from the rising phase of the onset curve, as the slope of a
quasi-Poisson GLM (quasi-Poisson because the daily counts are badly overdispersed).

```{r}
growth <- inc40 |>
  filter(between(time_of_symptom_onset, 8, 26)) |>
  count(time_of_symptom_onset)

m <- glm(n ~ time_of_symptom_onset, quasipoisson, growth)
r <- coef(m)[[2]]
r_se <- sqrt(vcov(m)[2, 2])
sprintf("growth rate r = %.2f /day (95%% CI %.2f, %.2f)", r, r - 1.96 * r_se, r + 1.96 * r_se)
```

```{r}
set.seed(1)
draws <- list(
  "backward (observed)" = draw(fit(rep(1, nrow(gi)))),
  "forward (growth-weighted)" = draw(fit(exp(r * gi$tau_mid)))
)

fmt <- \(x) sprintf("%.1f (%.1f, %.1f)", median(x), quantile(x, .025), quantile(x, .975))
draws |>
  lapply(\(d) tibble::tibble(`mean (days)` = fmt(d$mean), `sd (days)` = fmt(d$sd))) |>
  bind_rows(.id = "estimand") |>
  knitr::kable()
```

```{r}
#| fig-width: 6
#| fig-height: 3.5
grid <- seq(0, 14, 0.1)

# 95% density band per estimand (gamma recovered from the mean/sd draws)
band <- draws |>
  lapply(\(d) {
    i <- sample(length(d$mean), 600)
    shape <- (d$mean[i] / d$sd[i])^2
    rate <- d$mean[i] / d$sd[i]^2
    dens <- mapply(\(sh, rt) dgamma(grid, sh, rt), shape, rate)
    tibble::tibble(
      t = grid,
      med = apply(dens, 1, median),
      lo = apply(dens, 1, quantile, .025),
      hi = apply(dens, 1, quantile, .975)
    )
  }) |>
  bind_rows(.id = "estimand")

ggplot(band, aes(t, med, colour = estimand, fill = estimand)) +
  geom_histogram(
    data = tibble::tibble(d = gi$tau_mid),
    aes(d, after_stat(density)),
    inherit.aes = FALSE, binwidth = 1, fill = "grey88", colour = "white"
  ) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 1) +
  labs(x = "generation interval (days)", y = "density", colour = NULL, fill = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The backward interval is about 3 days; undoing the tilt moves it to about 5. `r`
is tight enough (above) that propagating its uncertainty barely changes the
forward CI, so the band is the fit uncertainty at the point `r`, and most of the
width is the 68 pairs.

Two caveats. The forward estimate removes the population tilt but not the
within-household competition, where the recorded infector is whoever transmitted
first among several. That competition is small here: only about 6% of infectees
had two or more potential household infectors, since households reach at most
four people. The weighting also leans on the few long pairs, so read the forward
number as direction and scale.
